Carrier phase and amplitude estimation for phase shift keying using pilots and data

### 2013-01-07 paper.tex
added some of Vaughan's comments

\begin{abstract}
We consider least squares estimators of carrier phase and amplitude from a noisy communications signal that contains both pilot signals, known to the receiver, and data signals, unknown to the receiver.  We focus on signaling constellations that have symbols evenly distributed on the complex unit circle, i.e., $M$-ary phase shift keying.  We show, under reasonably mild conditions on the distribution of the noise, that the least squares estimator of carrier phase is strongly consistent and asymptotically normally distributed.  However, the amplitude estimator is not consistent, but converges to a positive real number that is a function of the true carrier amplitude, the noise distribution and the size of the constellation.  Our theoretical results can also be applied to the case where no pilot symbols exist, i.e., noncoherent detection.  The results of Monte Carlo simulations are provided and these agree with the theoretical results.   
\end{abstract}
\begin{IEEEkeywords}
Coherent detection, noncoherent detection, phase shift keying, asymptotic statistics
\end{IEEEkeywords}

@@

\begin{lemma}\label{lem:q1k1parts}
Put
\[
q_1(\lambda) + j k_1(\lambda) = \expect\big[ R_1 e^{j(\lambda + \Phi_1)} - R_1 e^{j\Phi_1}  \big].
\]
We have $q_1(\lambda) = \hat{\lambda}_L o_P(1)$ and $k_1(\hat{\lambda}_L) = \hat{\lambda}_L\big(1 + o_P(1) \big)$.
\end{lemma}
\begin{proof}
We have
\[
q_1(\lambda) + j k_1(\lambda) = (e^{j\lambda} - 1) \expect R_1e^{j\Phi_1} = e^{j\lambda} - 1
\]
since the mean of $R_1e^{j\Phi_1}$ is 1.  By a first order expansion about $\lambda = 0$ we obtain
\[
q_1(\lambda) + j k_1(\lambda) = \lambda\big( j + O(\lambda) \big)
\]
Since $\hat{\lambda}_L$ converges almost surely to zero as $L\rightarrow\infty$ it follows that $O(\hat{\lambda}_L) = o_P(1)$.  Thus
\[
q_1(\hat{\lambda}_L) + j k_1(\hat{\lambda}_L) = \hat{\lambda}_L\big( j + o_P(1) \big)
\]
and the lemma follow by taking real and imaginary parts.
\end{proof}

\begin{lemma}\label{lem:q2k2parts}
Put
\[
q_2(\lambda) + j k_2(\lambda) = \expect\big[ R_1 e^{j\fracpart{\lambda + \Phi_1}} - R_1 e^{j\fracpart{\Phi_1}}  \big].
\]
We have $q_2(\hat{\lambda}_L) = \hat{\lambda}_L o_P(1)$ and $k_2(\hat{\lambda}_L) = \hat{\lambda}_L \big( H + o_P(1) \big)$ where $H$ is defined in the statement of Theorem~\ref{thm:normality}.
\end{lemma}
\begin{proof}
Because $\hat{\lambda}_L \rightarrow 0$ almost surely as $L\rightarrow\infty$, it is only the behaviour of $q_2(\lambda)$ and $k_2(\lambda)$ around $\lambda = 0$ that is relevant.  We will examine $q_2(\lambda)$ and $k_2(\lambda)$ for $0 \leq \lambda < \tfrac{\pi}{M}$.  An analogous argument follows when $-\tfrac{\pi}{M} < \lambda < 0$.  To keep our notation clean put 
\[
\psi_k = \frac{2\pi}{M}k + \frac{\pi}{M}
\] 
with $k \in \ints$.  When $\Phi_1 \in [\psi_{k-1}, \psi_k - \lambda)$,
\begin{align*}
e^{j\fracpart{\lambda + \Phi_1}} - e^{j\fracpart{\Phi_1}} &=  e^{j\big(\lambda + \Phi_1 - \tfrac{2\pi}{M}k\big)} - e^{j\big( \Phi_1- \tfrac{2\pi}{M}k\big)} \\
&= \big(e^{j\lambda} - 1 \big) e^{j\big(\Phi_1- \tfrac{2\pi}{M}k\big)} \\
&= \big(e^{j\lambda} - 1 \big) e^{j\fracpart{\Phi_1}},
\end{align*}
and when $\Phi_1 \in [\psi_{k} - \lambda, \psi_k)$,
\begin{align*}
e^{j\fracpart{\lambda + \Phi_1}} - &e^{j\fracpart{\Phi_1}} \\
&=  e^{j\big(\lambda + \Phi_1 - \tfrac{2\pi}{M}k - \tfrac{2\pi}{M}\big)} - e^{j\big( \Phi_1- \tfrac{2\pi}{M}k\big)} \\
&= \big(e^{j\big(\lambda - \tfrac{2\pi}{M}\big)} - 1 \big) e^{j\fracpart{\Phi_1}} \\
&= \big(e^{j\lambda} - 1 \big) e^{j\fracpart{\Phi_1}}  + e^{j\lambda}\big( e^{-j\tfrac{2\pi}{M}} - 1 \big)e^{j\fracpart{\Phi_1}}.
\end{align*}
Thus, when $\Phi_1 \in [\psi_{k-1}, \psi_k)$,
\begin{align*}
e^{j\fracpart{\lambda + \Phi_1}} - e^{j\fracpart{\Phi_1}} &= \big(e^{j\lambda} - 1 \big) e^{j\fracpart{\Phi_1}} \\
&\hspace{0.5cm} + e^{j\lambda}\big( e^{-j\tfrac{2\pi}{M}} - 1 \big)e^{j\fracpart{\Phi_1}} \chi_k(\Phi_1,\lambda)
\end{align*}
where 
\[
\chi_k(\Phi_1,\lambda) = \begin{cases}
1, & \Phi_1 \in [\psi_{k} - \lambda, \psi_k)  \\
0, & \text{otherwise}.
\end{cases}
\]
Now
\begin{align}
q_2(\lambda) + j k_2(\lambda) &= \expect\big[ R_1 e^{j\fracpart{\lambda + \Phi_1}} - R_1 e^{j\fracpart{\Phi_1}}  \big] \nonumber \\
&= \big(e^{j\lambda} - 1 \big) h_2(0)  + e^{j\lambda} B(\lambda) \label{eq:q2k2withB}
\end{align}
since $\expect R_1 e^{j\fracpart{\Phi_1}} = h_2(0) + \expect R_1 \sin\fracpart{\Phi_1} = h_2(0)$ as a result of Lemma~\ref{lem:expectImRfracpart} and where
\begin{equation}\label{eq:Blambdadefn}
B(\lambda) = \big( e^{-j\tfrac{2\pi}{M}} - 1 \big)\expect R_1 e^{j\fracpart{\Phi_1}} \chi(\Phi_1,\lambda)
\end{equation}
and $\chi(\Phi_1, \lambda) = \sum_{k \in \ints}\chi_k(\Phi_1, \lambda)$.

Now $e^{j\hat{\lambda}_L} = 1 + o_P(1)$ and $e^{j\hat{\lambda}_L} - 1 = \hat{\lambda}_L \big( j  + o_P(1) \big)$ by the argument in Lemma~\ref{lem:q1k1parts} and
\[
B(\hat{\lambda}_L) = -\hat{\lambda}_L\left(2j\sin(\tfrac{\pi}{M}) \sum_{k=0}^{M-1} g(\psi_k) + o_P(1)\right) 
\]
by Lemma~\ref{lem:Blambdaconv}.  Combining these results into~\eqref{eq:q2k2withB} we obtain
\begin{align*}
q_2(\lambda) + &jk_2(\hat{\lambda}_L) \\
&= \hat{\lambda}_L \left( jh_2(0) - 2j\sin(\tfrac{\pi}{M}) \sum_{k=0}^{M-1} g(\psi_k) + o_P(1) \right) \\
&= \hat{\lambda}_L\big( jH + o_P(1) \big)
\end{align*}
and the lemma follows by taking real and imaginary parts.
\end{proof}

\begin{lemma}\label{lem:Blambdaconv}
With $B(\lambda)$ defined in~\eqref{eq:Blambdadefn} we have 
\[
B(\hat{\lambda}_L) = -\hat{\lambda}_L\left(2j\sin(\tfrac{\pi}{M}) \sum_{k=0}^{M-1} g(\psi_k) + o_P(1)\right) 
\]
\end{lemma}
\begin{proof}
Put $A(\lambda) = \expect R_1 e^{j\fracpart{\Phi_1}} \chi(\Phi_1,\lambda)$.  Recalling that $f(r,\phi)$ is the joint pdf of $R_1$ and $\Phi_1$ we have
\begin{align*}
 A(\lambda)  &= \int_{0}^{2\pi}\int_{0}^{\infty} r f(r,\phi) e^{j\fracpart{\phi}}\chi(\phi,\lambda) dr d\phi \\
&= \sum_{k\in\ints}\int_{0}^{2\pi} g(\phi) e^{j\fracpart{\phi}} \chi_k(\phi,\lambda) d\phi \\
&= \sum_{k=0}^{M-1}\int_{\psi_k-\lambda}^{\psi_k} g(\phi)  e^{j\fracpart{\phi}} d\phi,
\end{align*}
the last line because the $\chi_k(\phi,\lambda)$ terms inside the integral are zero for all $\phi \in [0,2\pi]$ when $k \notin \{0,\dots,M-1\}$.  Observe that $\fracpart{\phi} \rightarrow \tfrac{\pi}{M}$ as $\phi$ approaches $\psi_k$ from below.  Because $g(\psi_k)$ is continuous at $\psi_k$ for each $k = 0, \dots, M-1$ (by assumption in Theorem~\ref{thm:normality}) we have 
\[
\frac{1}{\lambda} A(\lambda) \rightarrow \sum_{k=0}^{M-1} g(\psi_k) e^{j\tfrac{\pi}{M}}
\]
as $\lambda$ approaches zero from above.  We are only interested in the limit from above because we are working under the assumption that $0 \leq \lambda < \frac{\pi}{M}$ (see the proof of Lemma~\ref{lem:q2k2parts}).  The analogous argument when $-\tfrac{\pi}{M} \leq \lambda < 0$ would involve limits as $\lambda$ approaches zero from below.  Thus
\[
A(\hat{\lambda}_L) = \hat{\lambda}_L \left(  e^{j\tfrac{\pi}{M}} \sum_{k=0}^{M-1} g(\psi_k) + o_P(1) \right)
\]
and the lemma follows since $\big( e^{-j\tfrac{2\pi}{M}} - 1 \big) e^{j\frac{\pi}{M}} = -2j\sin(\tfrac{\pi}{M})$ and $B(\lambda) = \big( e^{-j\tfrac{2\pi}{M}} - 1 \big)A(\lambda)$.
\end{proof}

>>>>

\begin{abstract}
We consider least squares estimators of carrier phase and amplitude from a noisy communications signal that contains both pilot signals, known to the receiver, and data signals, unknown to the receiver.  We focus on signaling constellations that have symbols evenly distributed on the complex unit circle, i.e., $M$-ary phase shift keying.  We show, under reasonably mild conditions on the distribution of the noise, that the least squares estimator of carrier phase is strongly consistent and asymptotically normally distributed.  However, the amplitude estimator is not consistent, but converges to a positive real number that is a function of the true carrier amplitude, the noise distribution and the size of the constellation.  Our theoretical results can also be applied to the case where no pilot symbols exist, i.e., noncoherent detection.  The results of Monte Carlo simulations are provided and these agree with the theoretical results.   
\end{abstract}
\begin{IEEEkeywords}
Detection, phase shift keying, asymptotic statistics
\end{IEEEkeywords}

@@

\begin{lemma}\label{lem:q1k1parts}
Put
\[
q_1(\lambda) + j k_1(\lambda) = \expect\big[ R_1 e^{j(\lambda + \Phi_1)} - R_1 e^{j\Phi_1}  \big].
\]
We have $q_1(\lambda) = \hat{\lambda}_L o_P(1)$ and $k_1(\hat{\lambda}_L) = \hat{\lambda}_L\big(1 + o_P(1) \big)$.
\end{lemma}
\begin{IEEEproof}
We have
\[
q_1(\lambda) + j k_1(\lambda) = (e^{j\lambda} - 1) \expect R_1e^{j\Phi_1} = e^{j\lambda} - 1
\]
since the mean of $R_1e^{j\Phi_1}$ is 1.  By a first order expansion about $\lambda = 0$ we obtain
\[
q_1(\lambda) + j k_1(\lambda) = \lambda\big( j + O(\lambda) \big)
\]
Since $\hat{\lambda}_L$ converges almost surely to zero as $L\rightarrow\infty$ it follows that $O(\hat{\lambda}_L) = o_P(1)$.  Thus
\[
q_1(\hat{\lambda}_L) + j k_1(\hat{\lambda}_L) = \hat{\lambda}_L\big( j + o_P(1) \big)
\]
and the lemma follow by taking real and imaginary parts.
\end{IEEEproof}

\begin{lemma}\label{lem:q2k2parts}
Put
\[
q_2(\lambda) + j k_2(\lambda) = \expect\big[ R_1 e^{j\fracpart{\lambda + \Phi_1}} - R_1 e^{j\fracpart{\Phi_1}}  \big].
\]
We have $q_2(\hat{\lambda}_L) = \hat{\lambda}_L o_P(1)$ and $k_2(\hat{\lambda}_L) = \hat{\lambda}_L \big( H + o_P(1) \big)$ where $H$ is defined in the statement of Theorem~\ref{thm:normality}.
\end{lemma}
\begin{IEEEproof}
Because $\hat{\lambda}_L \rightarrow 0$ almost surely as $L\rightarrow\infty$, it is only the behaviour of $q_2(\lambda)$ and $k_2(\lambda)$ around $\lambda = 0$ that is relevant.  We will examine $q_2(\lambda)$ and $k_2(\lambda)$ for $0 \leq \lambda < \tfrac{\pi}{M}$.  An analogous argument follows when $-\tfrac{\pi}{M} < \lambda < 0$.  To keep our notation clean put 
\[
\psi_k = \frac{2\pi}{M}k + \frac{\pi}{M}
\] 
with $k \in \ints$.  When $\Phi_1 \in [\psi_{k-1}, \psi_k - \lambda)$,
\begin{align*}
e^{j\fracpart{\lambda + \Phi_1}} - e^{j\fracpart{\Phi_1}} &=  e^{j\big(\lambda + \Phi_1 - \tfrac{2\pi}{M}k\big)} - e^{j\big( \Phi_1- \tfrac{2\pi}{M}k\big)} \\
&= \big(e^{j\lambda} - 1 \big) e^{j\big(\Phi_1- \tfrac{2\pi}{M}k\big)} \\
&= \big(e^{j\lambda} - 1 \big) e^{j\fracpart{\Phi_1}},
\end{align*}
and when $\Phi_1 \in [\psi_{k} - \lambda, \psi_k)$,
\begin{align*}
e^{j\fracpart{\lambda + \Phi_1}} - &e^{j\fracpart{\Phi_1}} \\
&=  e^{j\big(\lambda + \Phi_1 - \tfrac{2\pi}{M}k - \tfrac{2\pi}{M}\big)} - e^{j\big( \Phi_1- \tfrac{2\pi}{M}k\big)} \\
&= \big(e^{j\big(\lambda - \tfrac{2\pi}{M}\big)} - 1 \big) e^{j\fracpart{\Phi_1}} \\
&= \big(e^{j\lambda} - 1 \big) e^{j\fracpart{\Phi_1}}  + e^{j\lambda}\big( e^{-j\tfrac{2\pi}{M}} - 1 \big)e^{j\fracpart{\Phi_1}}.
\end{align*}
Thus, when $\Phi_1 \in [\psi_{k-1}, \psi_k)$,
\begin{align*}
e^{j\fracpart{\lambda + \Phi_1}} - e^{j\fracpart{\Phi_1}} &= \big(e^{j\lambda} - 1 \big) e^{j\fracpart{\Phi_1}} \\
&\hspace{0.5cm} + e^{j\lambda}\big( e^{-j\tfrac{2\pi}{M}} - 1 \big)e^{j\fracpart{\Phi_1}} \chi_k(\Phi_1,\lambda)
\end{align*}
where 
\[
\chi_k(\Phi_1,\lambda) = \begin{cases}
1, & \Phi_1 \in [\psi_{k} - \lambda, \psi_k)  \\
0, & \text{otherwise}.
\end{cases}
\]
Now
\begin{align}
q_2(\lambda) + j k_2(\lambda) &= \expect\big[ R_1 e^{j\fracpart{\lambda + \Phi_1}} - R_1 e^{j\fracpart{\Phi_1}}  \big] \nonumber \\
&= \big(e^{j\lambda} - 1 \big) h_2(0)  + e^{j\lambda} B(\lambda) \label{eq:q2k2withB}
\end{align}
since $\expect R_1 e^{j\fracpart{\Phi_1}} = h_2(0) + \expect R_1 \sin\fracpart{\Phi_1} = h_2(0)$ as a result of Lemma~\ref{lem:expectImRfracpart} and where
\begin{equation}\label{eq:Blambdadefn}
B(\lambda) = \big( e^{-j\tfrac{2\pi}{M}} - 1 \big)\expect R_1 e^{j\fracpart{\Phi_1}} \chi(\Phi_1,\lambda)
\end{equation}
and $\chi(\Phi_1, \lambda) = \sum_{k \in \ints}\chi_k(\Phi_1, \lambda)$.

Now $e^{j\hat{\lambda}_L} = 1 + o_P(1)$ and $e^{j\hat{\lambda}_L} - 1 = \hat{\lambda}_L \big( j  + o_P(1) \big)$ by the argument in Lemma~\ref{lem:q1k1parts} and
\[
B(\hat{\lambda}_L) = -\hat{\lambda}_L\left(2j\sin(\tfrac{\pi}{M}) \sum_{k=0}^{M-1} g(\psi_k) + o_P(1)\right) 
\]
by Lemma~\ref{lem:Blambdaconv}.  Combining these results into~\eqref{eq:q2k2withB} we obtain
\begin{align*}
q_2(\lambda) + &jk_2(\hat{\lambda}_L) \\
&= \hat{\lambda}_L \left( jh_2(0) - 2j\sin(\tfrac{\pi}{M}) \sum_{k=0}^{M-1} g(\psi_k) + o_P(1) \right) \\
&= \hat{\lambda}_L\big( jH + o_P(1) \big)
\end{align*}
and the lemma follows by taking real and imaginary parts.
\end{IEEEproof}

\begin{lemma}\label{lem:Blambdaconv}
With $B(\lambda)$ defined in~\eqref{eq:Blambdadefn} we have 
\[
B(\hat{\lambda}_L) = -\hat{\lambda}_L\left(2j\sin(\tfrac{\pi}{M}) \sum_{k=0}^{M-1} g(\psi_k) + o_P(1)\right) 
\]
\end{lemma}
\begin{IEEEproof}
Put $A(\lambda) = \expect R_1 e^{j\fracpart{\Phi_1}} \chi(\Phi_1,\lambda)$.  Recalling that $f(r,\phi)$ is the joint pdf of $R_1$ and $\Phi_1$ we have
\begin{align*}
 A(\lambda)  &= \int_{0}^{2\pi}\int_{0}^{\infty} r f(r,\phi) e^{j\fracpart{\phi}}\chi(\phi,\lambda) dr d\phi \\
&= \sum_{k\in\ints}\int_{0}^{2\pi} g(\phi) e^{j\fracpart{\phi}} \chi_k(\phi,\lambda) d\phi \\
&= \sum_{k=0}^{M-1}\int_{\psi_k-\lambda}^{\psi_k} g(\phi)  e^{j\fracpart{\phi}} d\phi,
\end{align*}
the last line because the $\chi_k(\phi,\lambda)$ terms inside the integral are zero for all $\phi \in [0,2\pi]$ when $k \notin \{0,\dots,M-1\}$.  Observe that $\fracpart{\phi} \rightarrow \tfrac{\pi}{M}$ as $\phi$ approaches $\psi_k$ from below.  Because $g(\psi_k)$ is continuous at $\psi_k$ for each $k = 0, \dots, M-1$ (by assumption in Theorem~\ref{thm:normality}) we have 
\[
\frac{1}{\lambda} A(\lambda) \rightarrow \sum_{k=0}^{M-1} g(\psi_k) e^{j\tfrac{\pi}{M}}
\]
as $\lambda$ approaches zero from above.  We are only interested in the limit from above because we are working under the assumption that $0 \leq \lambda < \frac{\pi}{M}$ (see the proof of Lemma~\ref{lem:q2k2parts}).  The analogous argument when $-\tfrac{\pi}{M} \leq \lambda < 0$ would involve limits as $\lambda$ approaches zero from below.  Thus
\[
A(\hat{\lambda}_L) = \hat{\lambda}_L \left(  e^{j\tfrac{\pi}{M}} \sum_{k=0}^{M-1} g(\psi_k) + o_P(1) \right)
\]
and the lemma follows since $\big( e^{-j\tfrac{2\pi}{M}} - 1 \big) e^{j\frac{\pi}{M}} = -2j\sin(\tfrac{\pi}{M})$ and $B(\lambda) = \big( e^{-j\tfrac{2\pi}{M}} - 1 \big)A(\lambda)$.
\end{IEEEproof}
